The role of cognateness in non-native spoken word recognition

### 2024-05-09 manuscript/manuscript.qmd
Add Methods section to Experiment 3

All materials, data, and code used in this study are hosted in the Open Science Framework [https://osf.io/9fjxm/](https://osf.io/9fjxm/) and a GitHub repository [https://github.com/gongcastro/translation-elicitation.git](https://github.com/gongcastro/translation-elicitation.git), along with additional notes.

@@

```{r tbl-participants}
#| label: tbl-participants
tbl_data <- participants |> 
    mutate(exp = case_when(
        source=="Experiment" & group!= "cat-SPA" ~ "Experiment 1",
        source=="Experiment" & group== "cat-SPA" ~ "Experiment 2",
        source=="Questionnaire" ~ "Experiment 3",
    )) |> 
    summarise(n = n(),
              n_excluded = sum(!valid_status=="Valid"),
              across(age, lst(mean, sd, min, max)),
              knows_l2 = sum(l2 != "None"),
              l2_lst = lst(unique(l2)[unique(l2)!="None"]),
              .by = c(group, exp))

tbl_data |> 
    gt(groupname_col = "exp", rowname_col = "group") |> 
    fmt_number(matches("age")) |> 
    fmt_integer(matches("min|max")) |> 
    cols_merge_range(age_min, age_max) |> 
    tab_spanner("Age", matches("age")) |> 
    tab_spanner("Sample size", starts_with("n")) |> 
    tab_spanner("L2", matches("l2")) |> 
    cols_label(group = "Group",
               exp = "Experiment",
               n = "N",
               n_excluded = "Excluded",
               age_mean = "M",
               age_sd = "SD",
               age_min = "Range",
               knows_l2 = "N knows L2?",
               l2_lst = "Which L2?") |> 
    tab_style(cell_text(style = "italic"),
              cells_column_labels()) |> 
    tab_style(cell_text(weight = "bold"),
              cells_column_spanners()) |> 
    tab_style(cell_text(align = "left"),
              list(cells_body(), cells_column_labels()))
```

@@

We arranged two lists of input words to be presented to participants in the auditory modality: one in Catalan and one in Spanish. Words in the Catalan list were `r round(lengths$cat_eng$ipa_flat_1_mean, 2)` phonemes long on average (*SD* = `r round(lengths$cat_eng$ipa_flat_1_sd, 2)`, *Range* = `r round(lengths$cat_eng$ipa_flat_1_min, 2)`-`r round(lengths$cat_eng$ipa_flat_1_max, 2)`), and the orthographic form of their English translations (which participants had to type), were `r round(lengths$cat_eng$word_2_mean, 2)` characters long on average (*SD* = `r round(lengths$cat_eng$word_2_sd, 2)`, *Range* = `r round(lengths$cat_eng$word_2_min, 2)`-`r round(lengths$cat_eng$word_2_max, 2)`. Words in the Spanish list were `r round(lengths$spa_eng$ipa_flat_1_mean, 2)` phonemes long on average (*SD* = `r round(lengths$spa_eng$ipa_flat_1_sd, 2)`, *Range* = `r round(lengths$spa_eng$ipa_flat_1_min, 2)`-`r round(lengths$spa_eng$ipa_flat_1_max, 2)`), and the orthographic form of their English translations (which participants had to type), were `r round(lengths$spa_eng$word_2_mean, 2)` characters long on average (*SD* = `r round(lengths$spa_eng$word_2_sd, 2)`, *Range* = `r round(lengths$spa_eng$word_2_min, 2)`-`r round(lengths$spa_eng$word_2_max, 2)`.

Participants listened to one audio file in each trial, each containing a single word presented in isolation. The audio files were the same ones used in child experiments conducted in the Laboratori de Recerca en Infància of Universitat Pompeu Fabra (Barcelona, Spain). These audio files were recorded by a proficient Catalan-Spanish female bilingual from the Metropolitan Area of Barcelona in a child-directed manner. Catalan and Spanish words were recorded at 44,100 Hz in separate files in the same session, and then de-noised using Audacity and normalised at peak intensity using Praat [@broersma2021praat]. The average duration of the Catalan audio files was `r round(durations$cat_eng$duration_mean, 2)` (*SD* = `r round(durations$cat_eng$duration_sd, 2)`, *Range* = `r round(durations$cat_eng$duration_min, 2)`-`r round(durations$cat_eng$duration_max, 2)`). The average duration of the Catalan audio files was `r round(durations$spa_eng$duration_mean, 2)` (*SD* = `r round(durations$spa_eng$duration_sd, 2)`, *Range* = `r round(durations$spa_eng$duration_min, 2)`-`r round(durations$spa_eng$duration_max, 2)`).

```{r}
stim_exc_freq <- stimuli |> 
    mutate(is_na = is.na(freq_zipf_2)) |> 
    count(group, is_na) |> 
    filter(is_na)
stim_exc_freq <- split(stim_exc_freq, stim_exc_freq$group)
names(stim_exc_freq) <- janitor::make_clean_names(names(stim_exc_freq))

@@

For each translation pair, we defined three predictors of interest: the lexical frequency of the correct translation (*Frequency*), the phonological similarity between the presented (Catalan or Spanish) word and their correct English translation (*Cognateness*), and presented word's number of cross-language phonological neighbours (*CLPN*).

We included *Frequency* under the hypothesis that, keeping other predictors constant, participants would translate higher-frequency words more accurately and faster than lower-frequency words. Lexical frequencies of correct translations were extracted from SUBTLEX-UK [@van2014subtlex], and transformed to Zipf scores [@van2014subtlex]. Words in the stimuli list without a lexical frequency value were excluded from data analysis (`r stim_exc_freq$cat_eng$n` in the Catalan list, `r stim_exc_freq$spa_eng$n` in the Spanish list).

We calculated *Cognatenesss*, our main predictor of interest, by computing the Levenshtein similarity between the X-SAMPA transcriptions of each pair of translations using the `stringdist` R package [@van2014stringdist]. The Levenshtein distance computes the edit distance between two character strings (in this case, phoneme transcriptions) by counting the number of additions, deletions, and substitutions necessary to make both strings identical [@levenshtein1966binary]. We divided this edit distance by the number of characters included in the longest X-SAMPA transcription of the translation pair. This results in a proportion score, in which values closer to zero correspond to lower Levenshtein distances between phonological transcriptions (i.e., higher similarity), and values closer to 1 correspond to higher Levenshtein distances (i.e., lower similarity). This transformation also help account for the fact that the Levenshtein distance tends to increase with the length of the transcriptions. For interpretability, we subtracted this proportion from one, so that values closer to one correspond to higher similarity between phonological transcriptions, and values closer to zero correspond to lower similarity between phonological transcriptions. For example, the *table* (teɪbəl)-*mesa* (mesa) translation pair had a `r scales::percent(stringsim(enc2utf8("teɪbəl"), enc2utf8("mesa")))` similarity, while the *train* (trEIn)-*tren* (t`r knitr::asis_output("\U027E")`En) translation pair had a `r scales::percent(stringsim(enc2utf8("trEIn"), enc2utf8("tɾEn")))` similarity. Figure 1 summarises the lexical frequency, phonological neighbourhood density and phonological overlap of the words included in the Catalan and the Spanish lists.

For each Catalan and Spanish word, we calculated the number of *CLPN* by counting the number of English words with same or higher lexical frequency, and whose phonological transcription (in X-SAMPA format) different in up to one phoneme from that of the presented Catalan or Spanish word. Lexical frequencies and phonological transcriptions were extracted from the multilingual database CLEARPOND [@marian2012clearpond]^[[Phonological trnascriptions in CLEARPOND were generated from eSPEAK, [http://espeak.sourceforge.net/](http://espeak.sourceforge.net/)]].

```{r tbl-stimuli}
#| label: tbl-stimuli
stimuli |> 
    summarise(across(c(freq_zipf_2, lv, neigh_n_h),
                     lst(mean, sd, min, max), 
                     na.rm = TRUE),
              .by = c(group)) |> 
    gt(rowname_col = "group") |> 
    tab_spanner("Frequency", matches("freq")) |> 
    tab_spanner("Cognateness", matches("lv")) |> 
    tab_spanner("CLPN", matches("neigh")) |> 
    fmt_number(is.numeric) |> 
    fmt_integer(c(neigh_n_h_min,
                  neigh_n_h_max)) |> 
    cols_merge_uncert(freq_zipf_2_mean, freq_zipf_2_sd) |> 
    cols_merge_uncert(lv_mean, lv_sd) |> 
    cols_merge_uncert(neigh_n_h_mean, neigh_n_h_sd) |> 
    cols_merge_range(freq_zipf_2_min, freq_zipf_2_max) |> 
    cols_merge_range(lv_min, lv_max) |> 
    cols_merge_range(neigh_n_h_min, neigh_n_h_max) |> 
    cols_label(freq_zipf_2_mean = "Mean ± SD",
               freq_zipf_2_sd = "SD",
               freq_zipf_2_min = "Range",
               lv_mean = "Mean ± SD",
               lv_sd = "SD",
               lv_min = "Range",
               neigh_n_h_mean = "Mean ± SD",
               neigh_n_h_sd = "SD",
               neigh_n_h_min = "Range") |> 
    tab_style(cell_text(weight = "bold"),
              cells_column_spanners()) |> 
    tab_style(cell_text(style = "italic"),
              cells_column_labels())

@@

Each trial started with a yellow fixation point presented during one second on the centre of the screen over a black background. After one second, the audio started playing while the dot remained being displayed until the audio ended. Upon the offset of the fixation point and audio, participants were prompted to write their answer by a `>` symbol. Typed letters were displayed in the screen in real time to provide visual feed-back to participants. Participants were allowed to correct their answer. Then, participants pressed the `RETURN/ENTER` key to confirm their answer and start and new trial. We excluded trials where the participant's answer was not a valid word in the target language or no answer was given. Trials in which the response was mistyped by only one character were counted as correct (e.g., "pengiun" instead of "penguin"), as long as the response did not correspond to a distinct word. Trials in which participants took longer than 10 seconds to respond were also excluded. Participants contributed a total of `r nrow(exp_responses)` valid trials (`r nrow(exp_responses[exp_responses$group %in% c("cat-ENG", "cat-SPA"),])` in Catalan, `r nrow(exp_responses[exp_responses$group %in% c("spa-ENG"),])` in Spanish). The task took approximately 15 minutes to be completed.

```{r}
#| label: procedure-figure
#| echo: false
#| message: false
#| warning: false
#| out-width: "80%"
#| fig-cap: "Schematic representation of a trial in the experimental task."
knitr::include_graphics(here::here("img", "design.png"))
```

@@

@@

We collected data for a total of `r format(nrow(dataset$cat_eng) + nrow(dataset$spa_eng), big.mark = ",")` trials completed by `r n_participants$cat_eng + n_participants$spa_eng` unique participants. Of those trials, `r format(nrow(dataset$cat_eng), big.mark = ",")` were provided by `r n_participants$cat_eng` unique participants who listened to Catalan words, and `r format(nrow(dataset$spa_eng), big.mark = ",")` trials were provided by `r n_participants$spa_eng` unique participants who listened to Spanish words. We excluded trials in which participants did not enter any text (*n* = `r r_validity$cat_eng["blank"]`), in which a response in a language other than English was provided (e.g., `agua`, *n* = `r r_validity$cat_eng["language"]`), in which participants did not provide a whole word (e.g., `f`, *n* = `r r_validity$cat_eng["incomplete"]`), and in which participants added comments to the experimenter (e.g., `unsure`, (*n* = `r r_validity$cat_eng["other"]`). In addition, we excluded data from participants that self-rated their oral and/or written skills in Catalan and Spanish, or any other second language as four or higher in a five-point scale (*n* = `r excluded_lang`), were diagnosed with a language (*n* = `r excluded_impairment`), or did not contribute more than 80% of valid trials (*n* = `r excluded_ntrials`). After applying trial-level and participant-level inclusion criteria, the resulting dataset included `r format(nrow(dataset_clean$cat_eng) + nrow(dataset_clean$cat_eng), big.mark = ",")` trials provided by `r n_distinct(dataset_clean$cat_eng$participant_id) + n_distinct(dataset_clean$cat_eng$participant_id)` unique participants. Of those trials, `r format(nrow(dataset_clean$cat_eng), big.mark = ",")` were provided by `r n_distinct(dataset_clean$cat_eng$participant_id)` unique participants who listened to Catalan words, and `r format(nrow(dataset_clean$spa_eng), big.mark = ",")` trials were provided by `r n_distinct(dataset_clean$spa_eng$participant_id)` unique participants who listened to Spanish words. Responses given by English participants to Catalan presented words were `r round(lengths$cat_eng$len_mean, 2)` characters long on average (*Median* = `r round(lengths$cat_eng$len_median, 2)`, *SD* = `r round(lengths$cat_eng$len_sd, 2)`, Range = `r round(lengths$cat_eng$len_min, 2)`-`r round(lengths$cat_eng$len_max, 2)`), while their translations to Spanish responses were `r round(lengths$spa_eng$len_mean, 2)` characters long on average (*Median* = `r round(lengths$spa_eng$len_median, 2)`, *SD* = `r round(lengths$spa_eng$len_sd, 2)`, Range = `r round(lengths$spa_eng$len_min, 2)`-`r round(lengths$spa_eng$len_max, 2)`).

```{r tbl-results}
#| label: tbl-results
tbl_data <- list("Experiment 1" = dataset_1,
                 "Experiment 2" = dataset_2,
                 "Experiment 3" = dataset_3) |> 
    bind_rows(.id = "exp") |> 
    count(exp, group, response_type) |> 
    pivot_wider(names_from = response_type,
                values_from = n,
                names_repair = janitor::make_clean_names) |>
    relocate(exp, group, correct, typo, wrong, false_friend) |> 
    mutate(n_correct = rowSums(cbind(correct, typo)),
           n_incorrect = rowSums(cbind(wrong, false_friend)),
           n_total = rowSums(cbind(n_correct, n_incorrect)),
           across(correct:false_friend, 
                  \(x) x / n_total,
                  .names = "{col}_prop")) |> 
    select(exp, group, correct, typo, wrong, false_friend,
           matches("prop"))

tbl_data |> 
    gt(rowname_col = "group", groupname_col = "exp") |> 
    fmt_missing(everything(), everything(), "--") |> 
    tab_spanner("Correct responses", 
                c(correct, typo)) |> 
    tab_spanner("Incorrect responses", 
                c(wrong, false_friend)) |> 
    fmt_integer(c(correct, typo, wrong, false_friend)) |> 
    fmt_percent(is.double) |> 
    cols_merge_n_pct(correct, col_pct = correct_prop) |> 
    cols_merge_n_pct(typo, col_pct = typo_prop) |> 
    cols_merge_n_pct(wrong, col_pct = wrong_prop) |> 
    cols_merge_n_pct(false_friend, col_pct = false_friend_prop) |>

    cols_label(correct = "Correct",
               typo = "Typo",
               wrong = "Wrong translation",
               false_friend = "False friend") |> 
    tab_style(cell_text(weight = "bold"), 
              cells_column_spanners()) |> 
    tab_style(cell_text(style = "italic"), 
              cells_column_labels())

@@

```{r}
c1 <- gather_draws(exp_1_m0, `b_.*`, `sd_.*`, regex = TRUE) |> 
    summarise(across(.value, lst(.median = median, 
                                 .lower = \(x) hdi(x)[1],
                                 .upper = \(x) hdi(x)[2],
                                 .rope = \(x) mean(between(x, -0.1, 0.1))),
                     .names = "{fn}"))
c1 <- split(c1, c1$.variable)
names(c1) <- janitor::make_clean_names(names(c1)) 
```

@@

tbl_data |> 
    gt(rowname_col = "group",
       groupname_col = "exp") |> 
    fmt_number(is.numeric, decimals = 2) |> 
    #gtExtras::gt_plt_dist(correct_list, type = "density") |> 
    fmt_integer(c(matches("sum|min|max"), n_participants)) |> 
    fmt_number(matches("correct"), scale_by = 100) |> 
    tab_spanner("Accuracy (%)", matches("correct")) |> 
    tab_spanner("Valid trials", matches("n_trials")) |> 
    cols_merge_uncert(n_trials_mean, n_trials_sd) |> 
    cols_merge_range(n_trials_min, n_trials_max) |> 
    cols_merge_uncert(n_trials_mean, n_trials_sd) |> 
    cols_merge_uncert(correct_mean, correct_sd) |> 
    cols_merge_range(correct_min, correct_max) |> 
    cols_label(n_trials_sum = "N trials",
               n_participants = "N",
               n_trials_mean = "Mean ± SD",
               n_trials_min = "Range",
               correct_mean = "Mean ± SD",
               correct_sem = "SE",
               correct_min = "Range") |> 
    tab_style(cell_text(style = "italic"),
              cells_column_labels()) |> 
    tab_style(cell_text(weight = "bold"),
              cells_column_spanners()) |> 
    tab_style(cell_text(align = "left"),
              list(cells_body(), cells_column_labels()))
```

```{r fig-epreds-1}
#| label: fig-epreds-1
#| fig-height: 6
#| fig-width: 10
get_epreds <- function(model, data, n = 100, 
                       lv = seq(0, 1, length.out = 100), 
                       neigh_n_h = c(0, 2, 4, 8, 12),
                       ...) {

>>>>

All materials, data, and code used in this study are hosted in the Open Science Framework [https://osf.io/9fjxm/](https://osf.io/9fjxm/?view_only=aab7636ce1af48cf832596a7ea9101c5/) and a GitHub repository [https://github.com/gongcastro/translation-elicitation.git](https://github.com/gongcastro/translation-elicitation.git), along with additional notes.

@@

```{r tbl-participants}
#| label: tbl-participants
tbl_data <- participants |> 
    mutate(exp = case_when(
        source=="Experiment" & group!= "cat-SPA" ~ "Experiment 1",
        source=="Experiment" & group== "cat-SPA" ~ "Experiment 2",
        source=="Questionnaire" ~ "Experiment 3",
    )) |> 
    summarise(n = n(),
              n_excluded = sum(!valid_status=="Valid"),
              across(age, lst(mean, sd, min, max)),
              l2_lst = list(l2),
              .by = c(group, exp)) |> 
    mutate(l2_lst = map(l2_lst, 
                        function(x) {
                            y <- table(x)
                            y <- y[names(y) != "None"]
                            y <- paste0(names(y), " (", y, ")")
                            return(y)
                        }),
           n_excluded = paste0(n, " (", n_excluded, ")")) |> 
    select(-n)

tbl_data |> 
    gt(groupname_col = "exp", rowname_col = "group") |> 
    fmt_number(matches("age")) |> 
    fmt_integer(matches("min|max")) |> 
    cols_merge_uncert(age_mean, age_sd) |> 
    cols_merge_range(age_min, age_max) |> 
    tab_spanner("Age", matches("age")) |> 
    cols_label(group = "Group",
               exp = "Experiment",
               n_excluded = "N",
               age_mean = "Mean ± SD ",
               age_min = "Range",
               l2_lst = "L2") |> 
    tab_style(cell_text(style = "italic"),
              cells_column_labels()) |> 
    tab_style(cell_text(weight = "bold"),
              cells_column_spanners()) |> 
    tab_style(cell_text(align = "left"),
              list(cells_body(), cells_column_labels())) |> 
    tab_style(cell_text(weight = "bold",
                        align = "center",
                        style = "normal"),
              cells_column_labels(c(l2_lst, n_excluded))) |> 
    tab_footnote(
        "Number of included participants (number of excluded participants.)", 
        locations = cells_column_labels(n_excluded))
```

@@

We arranged two lists of input words to be presented to participants in the auditory modality: one in Catalan and one in Spanish. Words in the Catalan list were `r round(lengths$cat_eng$ipa_flat_1_mean, 2)` phonemes long on average (*SD* = `r round(lengths$cat_eng$ipa_flat_1_sd, 2)`, Range = `r round(lengths$cat_eng$ipa_flat_1_min, 2)`-`r round(lengths$cat_eng$ipa_flat_1_max, 2)`), and the orthographic forms of their English translations (which participants had to type) were `r round(lengths$cat_eng$word_2_mean, 2)` characters long on average (*SD* = `r round(lengths$cat_eng$word_2_sd, 2)`, Range = `r round(lengths$cat_eng$word_2_min, 2)`-`r round(lengths$cat_eng$word_2_max, 2)`. Words in the Spanish list were `r round(lengths$spa_eng$ipa_flat_1_mean, 2)` phonemes long on average (*SD* = `r round(lengths$spa_eng$ipa_flat_1_sd, 2)`, Range = `r round(lengths$spa_eng$ipa_flat_1_min, 2)`-`r round(lengths$spa_eng$ipa_flat_1_max, 2)`), and the orthographic form of their English translations (which participants had to type), were `r round(lengths$spa_eng$word_2_mean, 2)` characters long on average (*SD* = `r round(lengths$spa_eng$word_2_sd, 2)`, Range = `r round(lengths$spa_eng$word_2_min, 2)`-`r round(lengths$spa_eng$word_2_max, 2)`.

Participants listened to one audio file in each trial, each containing a single word presented in isolation. The audio files were the same ones used in child experiments conducted in the Laboratori de Recerca en Infància of Universitat Pompeu Fabra (Barcelona, Spain). These audio files were recorded by a proficient Catalan-Spanish female bilingual from the Metropolitan Area of Barcelona in a child-directed manner. Catalan and Spanish words were recorded at 44,100 Hz in separate files in the same session, and then de-noised using Audacity and normalised at peak intensity using Praat [@broersma2021praat]. The average duration of the Catalan audio files was `r round(durations$cat_eng$duration_mean, 2)` (*SD* = `r round(durations$cat_eng$duration_sd, 2)`, Range = `r round(durations$cat_eng$duration_min, 2)`-`r round(durations$cat_eng$duration_max, 2)`). The average duration of the Catalan audio files was `r round(durations$spa_eng$duration_mean, 2)` (*SD* = `r round(durations$spa_eng$duration_sd, 2)`, Range = `r round(durations$spa_eng$duration_min, 2)`-`r round(durations$spa_eng$duration_max, 2)`).

```{r exc-stimuli}
stim_exc_freq <- stimuli |> 
    mutate(is_na = is.na(freq_zipf_2)) |> 
    count(group, is_na) |> 
    filter(is_na)
stim_exc_freq <- split(stim_exc_freq, stim_exc_freq$group)
names(stim_exc_freq) <- janitor::make_clean_names(names(stim_exc_freq))

@@

For each translation pair, we defined three predictors of interest: the lexical frequency of the correct translation (*Frequency*), the phonological similarity between the presented (Catalan or Spanish) word and their correct English translation (*Cognateness*), and the presented word's number of cross-language phonological neighbours (*CLPN*).

We included *Frequency* as a nuance predictor, under the hypothesis that---keeping other predictors constant---participants would translate higher-frequency words more accurately and faster than lower-frequency words. Lexical frequencies of correct translations were extracted from SUBTLEX-UK [@van2014subtlex], and transformed to Zipf scores. Words in the stimuli list without a lexical frequency value were excluded from data analysis (`r stim_exc_freq$cat_eng$n` in the Catalan list, `r stim_exc_freq$spa_eng$n` in the Spanish list).

We calculated *Cognatenesss*, our main predictor of interest, by computing the Levenshtein similarity between the X-SAMPA transcriptions of each pair of translations using the `stringdist` R package [@van2014stringdist]. The Levenshtein distance computes the edit distance between two character strings---in this case, phoneme transcriptions---by counting the number of additions, deletions, and substitutions necessary to make both strings identical [@levenshtein1966binary]. We divided this edit distance by the number of characters included in the longest X-SAMPA transcription of the translation pair. This results in a proportion score, in which values closer to zero correspond to lower Levenshtein distances between phonological transcriptions (i.e., higher similarity), and values closer to 1 correspond to higher Levenshtein distances (i.e., lower similarity). This transformation also help account for the fact that the Levenshtein distance tends to increase with the length of the transcriptions. For interpretability, we subtracted this proportion from one, so that values closer to one correspond to higher similarity between phonological transcriptions, and values closer to zero correspond to lower similarity between phonological transcriptions. For example, the *table* (teɪbəl)-*mesa* (mesa) translation pair had a `r scales::percent(stringsim(enc2utf8("teɪbəl"), enc2utf8("mesa")))` similarity, while the *train* (`trEIn`)-*tren* (t`r knitr::asis_output("\U027E")`En) translation pair had a `r scales::percent(stringsim(enc2utf8("trEIn"), enc2utf8("tɾEn")))` similarity. Figure 1 summarises the lexical frequency, phonological neighbourhood density and phonological overlap of the words included in the Catalan and the Spanish lists.

For each Catalan and Spanish word, we calculated the number of *CLPN* by counting the number of English words with same or higher lexical frequency, and whose phonological transcription (in X-SAMPA format) different in up to one phoneme from that of the presented Catalan or Spanish word. Lexical frequencies and phonological transcriptions were extracted from the multilingual database CLEARPOND [@marian2012clearpond]^[[Phonological transcriptions in CLEARPOND were generated from eSPEAK, [http://espeak.sourceforge.net/](http://espeak.sourceforge.net/)]].

```{r tbl-stimuli}
#| label: tbl-stimuli
#| tbl-cap: "Stimuli properties."
stimuli |> 
    summarise(across(c(freq_zipf_2, lv, neigh_n_h),
                     lst(mean, sd, min, max), 
                     na.rm = TRUE),
              .by = c(group)) |> 
    gt(rowname_col = "group") |> 
    tab_spanner("Frequency", matches("freq")) |> 
    tab_spanner("Cognateness", matches("lv")) |> 
    tab_spanner("CLPN", matches("neigh")) |> 
    fmt_number(is.numeric) |> 
    fmt_integer(c(neigh_n_h_min,
                  neigh_n_h_max)) |> 
    cols_merge_uncert(freq_zipf_2_mean, freq_zipf_2_sd) |> 
    cols_merge_uncert(lv_mean, lv_sd) |> 
    cols_merge_uncert(neigh_n_h_mean, neigh_n_h_sd) |> 
    cols_merge_range(freq_zipf_2_min, freq_zipf_2_max) |> 
    cols_merge_range(lv_min, lv_max) |> 
    cols_merge_range(neigh_n_h_min, neigh_n_h_max) |> 
    cols_label(freq_zipf_2_mean = "Mean ± SD",
               freq_zipf_2_sd = "SD",
               freq_zipf_2_min = "Range",
               lv_mean = "Mean ± SD",
               lv_sd = "SD",
               lv_min = "Range",
               neigh_n_h_mean = "Mean ± SD",
               neigh_n_h_sd = "SD",
               neigh_n_h_min = "Range") |> 
    tab_style(cell_text(weight = "bold"),
              cells_column_spanners()) |> 
    tab_style(cell_text(style = "italic"),
              cells_column_labels()) |> 
    grand_summary_rows(columns = matches("mean|sd"),
                       fns = list(label = md("**Mean**"), 
                                  id = "totals",
                                  fn = "mean"),
                       fmt = ~fmt_number(., n_sigfig = 3, suffixing = TRUE))
```

@@

Each trial started with a yellow fixation point presented during one second on the centre of the screen over a black background. After one second, the audio started playing while the dot remained being displayed until the audio ended. Upon the offset of the fixation point and audio, participants were prompted to write their answer by a `>` symbol. Typed letters were displayed in the screen in real time to provide visual feed-back to participants. Participants were allowed to correct their answer. Then, participants pressed the `RETURN/ENTER` key to confirm their answer and start and new trial.

![Schematic representation of a trial in the experimental task](_assets/img/design.png)

@@

#### Data processing

After data collection, participants answers were manually coded into the following categories: *Correct*, *Typo*, *Wrong*, *False friend*, *Other*. Responses were coded as *Correct* if the provided string of characters was identical to the orthographic form of the correct translation. Trials were coded as *Typo* if the participant provided a string of characters only one edit distance (addition, deletion, or substitution) apart from the orthographic form of the correct translation (e.g., "pengiun" instead of "penguin"), as long as the response did not correspond to a distinct English word. Responses were coded as *False friend* if the participant provided a phonologically similar incorrect translation. Responses were coded as *Other* (see Data analysis section for more details). Both *Correct* and *Typo* responses were considered as correct, while *Wrong* and *False friend* responses were considered as incorrect. *Other* responses were excluded from data analysis.

Trials in which participants took longer than 10 seconds to respond were also excluded. Participants contributed a total of `r nrow(exp_responses)` valid trials (`r nrow(exp_responses[exp_responses$group %in% c("cat-ENG", "cat-SPA"),])` in Catalan, `r nrow(exp_responses[exp_responses$group %in% c("spa-ENG"),])` in Spanish). The task took approximately 15 minutes to be completed.

#### Modelling approach and statistical inference

@@

We collected data for a total of `r format(nrow(dataset$cat_eng) + nrow(dataset$spa_eng), big.mark = ",")` trials completed by `r n_participants$cat_eng + n_participants$spa_eng` unique participants. Of those trials, `r format(nrow(dataset$cat_eng), big.mark = ",")` were provided by `r n_participants$cat_eng` unique participants who listened to Catalan words, and `r format(nrow(dataset$spa_eng), big.mark = ",")` trials were provided by `r n_participants$spa_eng` unique participants who listened to Spanish words. We excluded trials in which participants did not enter any text (*n* = `r r_validity$cat_eng["blank"]`), in which a response in a language other than English was provided (e.g., `agua`, *n* = `r r_validity$cat_eng["language"]`), in which participants did not provide a whole word (e.g., `f`, *n* = `r r_validity$cat_eng["incomplete"]`), and in which participants added comments to the experimenter (e.g., `unsure`, (*n* = `r r_validity$cat_eng["other"]`). In addition, we excluded data from participants that self-rated their oral and/or written skills in Catalan and Spanish, or any other second language as four or higher in a five-point scale (*n* = `r excluded_lang`), were diagnosed with a language (*n* = `r excluded_impairment`), or did not contribute more than 80% of valid trials (*n* = `r excluded_ntrials`).

After applying trial-level and participant-level inclusion criteria, the resulting dataset included `r format(nrow(dataset_clean$cat_eng) + nrow(dataset_clean$cat_eng), big.mark = ",")` trials provided by `r n_distinct(dataset_clean$cat_eng$participant_id) + n_distinct(dataset_clean$cat_eng$participant_id)` unique participants. Of those trials, `r format(nrow(dataset_clean$cat_eng), big.mark = ",")` were provided by `r n_distinct(dataset_clean$cat_eng$participant_id)` unique participants who listened to Catalan words, and `r format(nrow(dataset_clean$spa_eng), big.mark = ",")` trials were provided by `r n_distinct(dataset_clean$spa_eng$participant_id)` unique participants who listened to Spanish words. Responses given by English participants to Catalan presented words were `r round(lengths$cat_eng$len_mean, 2)` characters long on average (*Median* = `r round(lengths$cat_eng$len_median, 2)`, *SD* = `r round(lengths$cat_eng$len_sd, 2)`, Range = `r round(lengths$cat_eng$len_min, 2)`-`r round(lengths$cat_eng$len_max, 2)`), while their translations to Spanish responses were `r round(lengths$spa_eng$len_mean, 2)` characters long on average (*Median* = `r round(lengths$spa_eng$len_median, 2)`, *SD* = `r round(lengths$spa_eng$len_sd, 2)`, Range = `r round(lengths$spa_eng$len_min, 2)`-`r round(lengths$spa_eng$len_max, 2)`).

```{r tbl-results}
#| label: tbl-results
tbl_data <- list("Experiment 1" = dataset_1,
                 "Experiment 2" = dataset_2,
                 "Experiment 3" = dataset_3) |> 
    bind_rows(.id = "exp") |> 
    count(exp, group, response_type) |> 
    pivot_wider(names_from = response_type,
                values_from = n,
                names_repair = janitor::make_clean_names) |>
    relocate(exp, group, correct, typo, wrong, false_friend) |> 
    mutate(n_correct = rowSums(cbind(correct, typo)),
           n_incorrect = rowSums(cbind(wrong, false_friend)),
           n_total = rowSums(cbind(n_correct, n_incorrect)),
           across(correct:false_friend, 
                  \(x) x / n_total,
                  .names = "{col}_prop")) |> 
    select(exp, group, correct, typo, wrong, false_friend,
           matches("prop")) |> 
    relocate(correct, correct_prop,
             typo, typo_prop,
             wrong, wrong_prop,
             false_friend, false_friend_prop)

tbl_data |> 
    gt(rowname_col = "group", groupname_col = "exp") |> 
    fmt_missing(everything(), everything(), "--") |> 
    tab_spanner("Correct responses", 
                matches("correct|typo")) |> 
    tab_spanner("Incorrect responses", 
                matches("wrong|false")) |> 
    fmt_integer(c(correct, typo, wrong, false_friend)) |> 
    fmt_number(is.double, pattern = "({x})", scale_by = 100) |> 
    cols_label(correct = "Correct",
               typo = "Typo",
               wrong = "Wrong",
               false_friend = "False friend",
               correct_prop = "(%)",
               typo_prop = "(%)",
               wrong_prop = "(%)",
               false_friend_prop = "(%)") |> 
    tab_style(cell_text(weight = "bold"), 
              cells_column_spanners()) |> 
    tab_style(cell_text(style = "italic"), 
              cells_column_labels()) |> 
    summary_rows(columns = is.integer,
                 fns = list(label = md("*Sum*"),
                            id = "totals",
                            fn = "sum"),
                 fmt = ~fmt_integer(.)) |>
    summary_rows(columns = matches("prop"),
                 fns = list(label = md("*Mean*"), 
                            id = "means",
                            fn = "mean"),
                 fmt = ~fmt_number(., scale_by = 100)) |> 
    tab_style(cell_text(align = "center"),
              cells_column_labels())

@@

```{r coefs-1}
c1 <- gather_draws(exp_1_m0, `b_.*`, `sd_.*`, regex = TRUE) |> 
    summarise(across(.value, lst(.median = median, 
                                 .lower = \(x) hdi(x)[1],
                                 .upper = \(x) hdi(x)[2],
                                 .rope = \(x) mean(between(x, -0.1, 0.1))),
                     .names = "{fn}"))
c1 <- split(c1, c1$.variable)
names(c1) <- janitor::make_clean_names(names(c1)) 
```

@@

tbl_data |> 
    gt(rowname_col = "group",
       groupname_col = "exp") |> 
    fmt_number(is.numeric, decimals = 2) |> 
    #gtExtras::gt_plt_dist(correct_list, type = "density") |> 
    fmt_integer(c(matches("sum|min|max"), n_participants)) |> 
    fmt_number(matches("correct"), scale_by = 100) |> 
    tab_spanner("Accuracy (%)", matches("correct")) |> 
    tab_spanner("Valid trials", matches("n_trials")) |> 
    cols_merge_range(n_trials_min, n_trials_max) |> 
    cols_merge_range(correct_min, correct_max) |> 
    cols_label(n_trials_sum = "N trials",
               n_participants = "N",
               n_trials_mean = "Mean",
               n_trials_sd = "SD",
               n_trials_min = "Range",
               correct_mean = "Mean",
               correct_sd = "SD",
               correct_sem = "SE",
               correct_min = "Range") |> 
    summary_rows(columns = is.integer,
                 fns = list(label = md("*Sum*"),
                            id = "totals",
                            fn = "sum"),
                 fmt = ~fmt_integer(.)) |>
    summary_rows(columns = matches("mean|sd|sem"),
                 fns = list(label = md("*Mean*"), 
                            id = "means",
                            fn = "mean"),
                 fmt = ~fmt_number(., scale_by = 100)) |> 
    tab_style(cell_text(align = "center"),
              cells_column_labels()) |> 
    tab_style(cell_text(style = "italic"),
              cells_column_labels()) |> 
    tab_style(cell_text(weight = "bold"),
              cells_column_spanners()) |> 
    tab_style(cell_text(align = "left"),
              list(cells_body(), cells_column_labels()))
```

```{r fig-epreds-1}
#| label: fig-epreds-1
#| fig-height: 6
#| fig-width: 10
#| fig-cap: "Posterior model-predicted mean accuracy in Experiment 1. Predictions were generated from 4,000 posterior samples, extracted for different values of *CLPN* (0, 2, 4, 8, 12) and *Cognateness* (1-100). Predictions are plotted separately for English participants translating Catalan words, and for English participants translating Spanish words. Lines indicate mean predictions, and intervals indicate 95%, 89%, 78%, 67%, and 50% credible intervals (CrI)."
get_epreds <- function(model, data, n = 1000, 
                       lv = seq(0, 1, length.out = 100), 
                       neigh_n_h = c(0, 2, 4, 8, 12),
                       ...) {
